Anxiety: a decision-theoretic perspective

### 2019-01-13 manuscript/manuscript.tex
sunday writing (pt 1)

@@

\author{
Samuel Zorowitz \\
Princeton Neuroscience Institute\\
Princeton University\\
Princeton, NJ 08540 \\
\texttt{zorowitz@princeton.edu} \\
\And
Nathaniel Daw \\
Princeton Neuroscience Institute\\
Princeton University\\
Princeton, NJ 08540 \\
\texttt{ndaw@princeton.edu} \\
}

@@

\section{Introduction}
Many psychiatric disorders involve distortions in models of the external world
and, correspondingly, aberrations of learning and decision making. As a normative
framework, decision theory allows us to describe and pose quantita­tive questions
about optimal and approximately optimal choice behavior \cite{DayanDaw2008} and
is, therefore, a critical tool for modeling, understanding, and predicting the
psychological and neurobiological mechanisms responsible for the abnormalities
observed in psychiatric behavior \cite{HuysDawDayan2015}. Many recent investigations
employing this approach have provided important insights into psychiatric symptoms,
including anhedonia in major depression \cite{Rutledge2017}, mood instability
in bipolar disorder \cite{EldarNiv2015, EldarDolanNiv2016},  habitual decision
making in disorders of compulsivity \cite{gillan2016}, and hallucinations in
schizophrenia \cite{powers2017, corlett2018}.

Anxiety disorders (generalized anxiety disorder, social anxiety disorder, panic
disorder, obsessive-compulsive disorder) represent the most common family of
psychiatric disorders (citation). Despite the differences in the content of worry,
all anxiety disorders share two core features: pessimistic inference and excessive
avoidance. Pessimistic inference describes a tendency to appraise threatening
situations as more frequent and/or more severe than they objectively are. This has
been reliably observed experimentally, as in fear conditioning wherein individuals
with anxiety exhibit exaggerated fear towards safe cues (CS-) and extinguished
threatening cues \cite{lissek2005, Duits2015}, and clinically, through self-report
(pick 3 citaitons). Excessive avoidance describes the corresponding tendency for
individuals with anxiety to consistently flee or avoid threatening situations.
Avoidance is often chosen even at the expense of potential reward (approch-avoidance
conflict), and is frequently a major impediment to recovery \cite{Arnaudova2017}.

Recent studies have attempted to provide some computational models of this behavior
\cite{Moutoussis2008, Maia2010}. For example, \cite{Maia2010} was able to show
how reinforcement learning could explain sustained avoidance in the absence of
repeated reinforcement, observed in rodents \cite{servatius2008} and something
not explained by two-factor theory \cite{Krypotos2015}. As has been discussed
elsewhere \cite{Moutoussis2017}, such accounts only explain the maintenance of
avoidance in short timescales, but do not explain why resistant to change and
why they do not go away in general. Such theories also do not explain the magnitude
of pessimistic inference, or the fear generalization. A broader account is needed.

@@

Alternately, an agent may choose optimally in the present but may be uncertain of
its choice control in the future. In such a case, we deviate from the optimal
policy in the future. We can adopt a different function. There are a number to
choose from \cite{Garcia2015}. Here we adopt the beta-pessimism criterion
\cite{Gaskett2003}:

@@

First and foremost, pathological anxiety is associated with pessimistic expectations
about the probability and cost of future negative events (Grupe \& Nitschke, 2013;
Bishop \& Gagne, 2018). In fear conditioning, individuals with anxiety are more
likely to exhibit increased threat appraisal for safety signals (CS-) and the
extinguished fear cue (CS+) (Lissek et al. 2005; Duits et al., 2015). In considering
hypothetical future negative outcomes, individuals with anxiety are more likely
to evaluate negative outcomes as more likely and their consequences more severe
(pick a few references).

In the first series of simulations, we show directly how lack of perceived control
can directly account for pessimistic inference. In the open field environment
(Figure 1a), an agent navigates a simple grid-world. The world consists of almost
entirely of empty tiles, except for two which yield a reward and punishment,
respectively.

Under complete control (objective or perceived), an agent in this environment is
under no threat despite the presence of an aversive state. This is simply because,
in a full controllable world, a (non-masochistic) agent is never forced to interact
with the aversive state. When this control breaks down -- through a stochastic
world or a distrust of future competency -- an agent can no longer guarantee it
will never encounter the aversive state. In such a scenario, the negative value
back-propagates to its predecessor states through the recursive formulas defined
above. The result is an environment that appears worse (i.e. less positive) than
it otherwise would.

We shoe this above. Using value iteration. Show it for best-case scenario. Show
it for expected random. Show it for worst-case scenario.

@@

@@

1. Open with definition of avoidance. Follow \cite{Arnaudova2017} in definitions, i.e.
the basic idea that avoidance is the distancing of oneself from real or perceived
threat.
2. Describe cliff walking environment. Need to discuss that we are expanding
on the result originally described in \cite{Gaskett2003}.
3. With increasing pessimism, agent further distances self from cliff. Note that
this is the optimal policy under those set of beliefs about the world.
4. This can explain how avoidance is maintained. Even with luck and chance in the
the world, these sets of beliefs actually lead to and sustain these behavioral
policies.

@@

This can also be used to explain behavioral inhibition \cite{bach2015, khemka2017}.
The idea is slightly different than what is advocated. The claim here is that
in anxious individuals, the Q-values should be closer together, and thus there
should be longer waiting time. This requires demonstrating that decisions are
not made in advance; only at the time of decision making. This is something
I think we will have to argue. (Need to double check the literature here.)

@@

Basically reviewing the arguments of Huys here \cite{Huys2012, Lally2017}.
Need to get facts straight before going through all of this.

@@

Discuss previous computational approaches: This has also been previously been
investigated in previous computational modeling \cite{Huys2009}. There
the authors found that beliefs about a lack of controllability, mechanistically
modeled as corrupted priors about action-outcome contingencies, found deficient
learning in novel environments. The present results are no doubt related, but
address a different environment and a different behavior.

We discussed two mechanisms by which negative values can spread. There is a third
that we did not address here: that of the state itself. Here we assumed that states
were fully observable; in other words, there was no uncertainty about the state
an agent was currently occupying. In partially-observable Markov decision problems,
the state itself must be inferred from external and internal variables. This is
yet another means by which bad value can spread if undue likelihood is granted
to undesirable or aversive states. This ideas has previously been proposed to
explain some symptoms in anxiety \cite{Paulus2012}, and not without good reason.
Anxiety disorders involve negatively biased interpretation of ambiguous percepts
\cite{Hartley2012}, which may result in aberrant planning and increased avoidance
for similar reasons to those described above. This is an area for future research.

In this article, we have explored a computational principle of approach and
avoidance learning in sequential decision environments. We have not discussed,
however, the particular psychological or neurobiological mechanisms by which this
sort of learning may take place. This issue has been discussed in detail elsewhere
\cite{Bishop2018}, but we briefly discuss a few possibilities here. Mechanically,
we understand that complex decision problems are solved by mental processes that
exist on a spectrum between model-based and model-free (Daw et al., 2005). The
difference between these two families of methods hinges on the reliance of an
internal model of the world used to make predictions about expected long-run
outcomes of particular choices.

@@

Within the realm of model-free decision making, several studies have found evidence
suggesting increased learning rates for negative reward prediction errors in
individuals with elevated levels of anxiety \cite{Aylward, Huang2017, Harle2017
garrett2018}. Though an asymmetry in sensitivity to
positive and negative reward prediction errors bears superficial similarity, for
the reasons described above merely attending more to negative information does not
necessarily predict increased avoidance insofar that avoidance can successfully
isolate negative outcomes. One possibility, and topic of future research, is how
prior beliefs about the reward statistics of the current environment dictate
learning rates. A second possibility is prioritized replay (Russek et al., 2017;
Mattar and Daw, 2018). Offline replay of previous experiences (through hippocampal
mechanisms) are known to facilitate learning. It has been proposed that biased
replay could bias value estimates (Gagne et al., 2018). This is another exciting
area for future research.

@@

\bibliographystyle{authordate1}
\bibliography{citations}

>>>>

\usepackage{natbib}

@@

\author{
Samuel Zorowitz \\
Princeton Neuroscience Institute\\
Princeton University\\
Princeton, NJ 08540 \\
\texttt{zorowitz@princeton.edu} \\
\And
Ida Momennejad \\
columbia University\\
New York, NY 10027 \\
\texttt{ida.m@columbia.edu} \\
\And
Nathaniel Daw \\
Princeton Neuroscience Institute\\
Princeton University\\
Princeton, NJ 08540 \\
\texttt{ndaw@princeton.edu} \\
}

@@

\section{Introduction}
Many psychiatric disorders involve distortions in models of the external world
and, correspondingly, aberrations of learning and decision making. As a normative
framework, decision theory allows us to describe and pose quantita­tive questions
about optimal and approximately optimal choice behavior \citep{DayanDaw2008} and
is, therefore, a critical tool for modeling, understanding, and predicting the
psychological and neurobiological mechanisms responsible for the abnormalities
observed in psychiatric behavior \citep{HuysDawDayan2015}. Many recent investigations
employing this approach have provided important insights into psychiatric symptoms,
including anhedonia in major depression \citep{Rutledge2017}, mood instability
in bipolar disorder \citep{EldarNiv2015, EldarDolanNiv2016},  habitual decision
making in disorders of compulsivity \citep{gillan2016}, and hallucinations in
schizophrenia \citep{powers2017, corlett2018}.

Anxiety disorders (generalized anxiety disorder, social anxiety disorder, panic
disorder, obsessive-compulsive disorder) represent the most common family of
psychiatric disorders (citation). Despite the differences in the content of worry,
all anxiety disorders share two core features: pessimistic inference and excessive
avoidance. Pessimistic inference describes a tendency to appraise threatening
situations as more frequent and/or more severe than they objectively are. This has
been reliably observed experimentally, as in fear conditioning wherein individuals
with anxiety exhibit exaggerated fear towards safe cues (CS-) and extinguished
threatening cues \citep{lissek2005, Duits2015}, and clinically, through self-report
(pick 3 citaitons). Excessive avoidance describes the corresponding tendency for
individuals with anxiety to consistently flee or avoid threatening situations.
Avoidance is often chosen even at the expense of potential reward (approch-avoidance
conflict), and is frequently a major impediment to recovery \citep{Arnaudova2017}.

Recent studies have attempted to provide some computational models of this behavior
\citep{Moutoussis2008, Maia2010}. For example, \citep{Maia2010} was able to show
how reinforcement learning could explain sustained avoidance in the absence of
repeated reinforcement, observed in rodents \citep{servatius2008} and something
not explained by two-factor theory \citep{Krypotos2015}. As has been discussed
elsewhere \citep{Moutoussis2017}, such accounts only explain the maintenance of
avoidance in short timescales, but do not explain why resistant to change and
why they do not go away in general. Such theories also do not explain the magnitude
of pessimistic inference, or the fear generalization. A broader account is needed.

@@

Alternately, an agent may choose optimally in the present but may be uncertain of
its choice control in the future. In such a case, we deviate from the optimal
policy in the future. We can adopt a different function. There are a number to
choose from \citep{Garcia2015}. Here we adopt the beta-pessimism criterion
\citep{Gaskett2003}:

@@

One symptom central to all of the anxiety disorders is pessimistic inference,
or a tendency to appraise the likelihood and/or severity of threatening situations
out of proportion to their actual environmental contingencies \citep{dsm5, BeckClark1997,
ClarkBeck2011}. For example, an individual with specific phobia may vastly exaggerate
the probability of encountering their fear (e.g. a venomous spider) in their
everyday lives. The first studies of pessimistic inference in clinically anxious
populations involved individuals judging the perceived likelihood and severity
of different aversive situations, with the general finding that anxious individuals
rated these as more likely than non-clinical individuals \citep{ButlerMathews1983,
ButlerMathews1987, Foa1996, MacLeod1996, MacLeod1997, Luten1997, Stober1997,
Borkovec1999}; a finding that has been replicated more recently \citep{Maner2006,
Mitte2007, Miranda2007}.

Experimentally, aberrations in threat appraisal have been repeatedly observed
in a litany of associative learning studies. Common to all of these studies,
participants learn that a particular stimulus (CS+) predicts the occurrence
of a negative outcome (e.g. shock, loud noise, aversive image). In simple conditioning,
there is some evidence to suggest that individuals with anxiety exhibit elevated
levels of response to threatening stimuli \citep{lissek2005, MinekaOehlberg2008,
Duits2015}. Importantly, conditioning studies also suggest impaired safety learning
in anxiety, wherein a stimulus never paired with an aversive outcome (CS-) is also
feared \citep{Duits2015}. Other notable findings include impaired blocking,
conditioned discrimination (AX+/BX-) (citations), and conditioned inhibition
(A+, AB-) (citations).

% I think when you come back to edit this, you need to highlight these findings
% more in the context of what you are about to explain. Pessimistic inference
% is both making bad things worse and, most importantly, that non-bad things
% become bad. Although one question is how much to include the other less
% studied paradigms

@@

In our first result, we demonstrate one simple mechanism by which pessimistic
inference and exaggerated threat appraisals may arise. In the *open field* environment,
an agent is placed in a large grid world entirely empty except for two states,
leading to positive and negative outcomes, respectively. Under complete control
(objective or perceived), an agent in this environment in under no threat despite
the presence of negative states. This is simply because, in a fully controllable
world, an agent need not ever interact with the aversive state. In fact, an agent
can decide to walk straight up to the aversive state and, insofar it does not
choose to take one step further, will never be in any actual danger.

This is not so when deterministic control cannot be assumed. If an agent believes
it is not in full control of its environment, then it cannot rule out the possibility
that it may inadvertently encounter the negative state at some point in the future,
especially if it veers to close to the state. This is true irrespective of whether
the agent believes it cannot guarantee its own future competency or the benevolence
of the environment. Importantly, this belief need not match the true environmental
statistics; belief is enough.

The long-run state values under different beliefs are plotted in Figure 1. As
expected, under optimistic beliefs (Figure 1b) the entire environment sans negative
state takes on a positive value (close to the positive state itself). As this belief
breaks down, we observe a different pattern in the long-run state values. When
future occupancy of the negative state becomes a non-zero probability, the undesired
action transition exerts influence on action-outcome calculations, back-propagating,
and thereby infecting antecedent states. As the agent's belief becomes increasingly
pessimistic, the more the worst possible action exerts influence and the more
negative, on average, the state-space grows.

In the breakdown of perceived control then, negative states can grow worse (i.e.
more predictive of future negative outcome) and otherwise positive states can too.
This occurs as a simple result that if control not guaranteed, cannot guaranteed
bad things wont happen.

@@

1. Open with definition of avoidance. Follow \citep{Arnaudova2017} in definitions, i.e.
the basic idea that avoidance is the distancing of oneself from real or perceived
threat.
2. Describe cliff walking environment. Need to discuss that we are expanding
on the result originally described in \citep{Gaskett2003}.
3. With increasing pessimism, agent further distances self from cliff. Note that
this is the optimal policy under those set of beliefs about the world.
4. This can explain how avoidance is maintained. Even with luck and chance in the
the world, these sets of beliefs actually lead to and sustain these behavioral
policies.

@@

This can also be used to explain behavioral inhibition \citep{bach2015, khemka2017}.
The idea is slightly different than what is advocated. The claim here is that
in anxious individuals, the Q-values should be closer together, and thus there
should be longer waiting time. This requires demonstrating that decisions are
not made in advance; only at the time of decision making. This is something
I think we will have to argue. (Need to double check the literature here.)

@@

Basically reviewing the arguments of Huys here \citep{Huys2012, Lally2017}.
Need to get facts straight before going through all of this.

@@

Discuss previous computational approaches: This has also been previously been
investigated in previous computational modeling \citep{Huys2009}. There
the authors found that beliefs about a lack of controllability, mechanistically
modeled as corrupted priors about action-outcome contingencies, found deficient
learning in novel environments. The present results are no doubt related, but
address a different environment and a different behavior.

We discussed two mechanisms by which negative values can spread. There is a third
that we did not address here: that of the state itself. Here we assumed that states
were fully observable; in other words, there was no uncertainty about the state
an agent was currently occupying. In partially-observable Markov decision problems,
the state itself must be inferred from external and internal variables. This is
yet another means by which bad value can spread if undue likelihood is granted
to undesirable or aversive states. This ideas has previously been proposed to
explain some symptoms in anxiety \citep{Paulus2012}, and not without good reason.
Anxiety disorders involve negatively biased interpretation of ambiguous percepts
\citep{Hartley2012}, which may result in aberrant planning and increased avoidance
for similar reasons to those described above. This is an area for future research.

In this article, we have explored a computational principle of approach and
avoidance learning in sequential decision environments. We have not discussed,
however, the particular psychological or neurobiological mechanisms by which this
sort of learning may take place. This issue has been discussed in detail elsewhere
\citep{Bishop2018}, but we briefly discuss a few possibilities here. Mechanically,
we understand that complex decision problems are solved by mental processes that
exist on a spectrum between model-based and model-free (Daw et al., 2005). The
difference between these two families of methods hinges on the reliance of an
internal model of the world used to make predictions about expected long-run
outcomes of particular choices.

@@

Within the realm of model-free decision making, several studies have found evidence
suggesting increased learning rates for negative reward prediction errors in
individuals with elevated levels of anxiety \citep{Aylward, Huang2017, Harle2017
garrett2018}. Though an asymmetry in sensitivity to
positive and negative reward prediction errors bears superficial similarity, for
the reasons described above merely attending more to negative information does not
necessarily predict increased avoidance insofar that avoidance can successfully
isolate negative outcomes. One possibility, and topic of future research, is how
prior beliefs about the reward statistics of the current environment dictate
learning rates. A second possibility is prioritized replay (Russek et al., 2017;
Mattar and Daw, 2018). Offline replay of previous experiences (through hippocampal
mechanisms) are known to facilitate learning. It has been proposed that biased
replay could bias value estimates (Gagne et al., 2018). This is another exciting
area for future research.

@@

\bibliographystyle{vancouver-authoryear}
\bibliography{citations}
